---
title: "Heart Rate Analysis"
author: "Teague"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(lubridate)
library(cowplot)
library(kableExtra)
```

### <i>Phyllostomus hastatus </i> heart rate exploration

These data have beeen cleaned to filter out QI scores of 3 & 2, as well as heart rates lower than 100 or higher than 800. Let's see some basic statistics for each individual

```{r }
hr <- read_csv(paste0(here(), "/processedData/filteredHR.csv"))
hr$timestamp <- force_tz(hr$timestamp, tz = "America/Panama") #make sure that the time zone is set correctly. These data should be recorded in EST and were read in at UTC. force_tz resets the time zone. If with_tz were used it would re-project the time into EST from UTC (so subtract 5 hours).

tsum <- hr %>% group_by(tagID, date(timestamp)) %>%
  summarize(meanHR = round(mean(hr.bpm),0),
            medianHR = round(median(hr.bpm), 0),
            minHR = min(hr.bpm),
            maxHR = max(hr.bpm),
            meanTemp = round(mean(temp),2),
            medianTemp = round(median(temp),2),
            minTemp = min(temp),
            maxTemp = max(temp))
kable(tsum)

```




I'm going to kick out pdfs of every day for each bat.
```{r}
days <- unique(date(hr$timestamp))
bats <- unique(hr$tagID)

for (i in 1:length(days)){
  day = days[i]
  tmp <- hr %>% filter(date(timestamp) == day)
  for (j in 1: length(bats)){
    bat = bats[j]
    plotdat <- tmp %>% filter(tagID == bat)
      p.hr <- plotdat %>% 
        ggplot()+
        geom_rect(xmin = ymd_hms(paste(day, "00:00:00"), tz= "America/Panama"),
            xmax = ymd_hms(paste(day, "06:00:00"), tz= "America/Panama"), ymin=-Inf, ymax=Inf,
            alpha=0.5,fill="grey90")+
        geom_rect(xmin = ymd_hms(paste(day, "19:00:00"), tz= "America/Panama"),
            xmax = ymd_hms(paste(day, "24:00:00"), tz= "America/Panama"), ymin=-Inf, ymax=Inf,
            alpha=0.5,fill="grey90")+
        geom_line(aes(x = timestamp, y = hr.bpm))+
        ggtitle(paste(day, bat))+
        theme_cowplot()
      
      p.tb <- plotdat %>% 
        ggplot()+
        geom_rect(xmin = ymd_hms(paste(day, "00:00:00"), tz= "America/Panama"),
            xmax = ymd_hms(paste(day, "06:00:00"), tz= "America/Panama"), ymin=-Inf, ymax=Inf,
            alpha=0.5,fill="grey90")+
        geom_rect(xmin = ymd_hms(paste(day, "19:00:00"), tz= "America/Panama"),
            xmax = ymd_hms(paste(day, "24:00:00"), tz= "America/Panama"), ymin=-Inf, ymax=Inf,
            alpha=0.5,fill="grey90")+
        geom_line(aes(x = timestamp,  y = temp), color = "red")+
        theme_cowplot()
      
      ggsave(paste0(here(), "/output/", bat, day, ".pdf"), plot_grid(p.hr, p.tb, ncol = 1), 
             width = 11, height = 6, units = "in")
  }
}

p.hr <- hr %>% filter(date(timestamp) == day & tagID == bats) %>% 
  ggplot()+
  geom_rect(xmin = ymd_hms(paste(day, "00:00:00"), tz= "America/Panama"),
            xmax = ymd_hms(paste(day, "06:00:00"), tz= "America/Panama"), ymin=-Inf, ymax=Inf, alpha=0.5,fill="grey90")+
  geom_rect(xmin = ymd_hms(paste(day, "19:00:00"), tz= "America/Panama"),
            xmax = ymd_hms(paste(day, "24:00:00"), tz= "America/Panama"), ymin=-Inf, ymax=Inf, alpha=0.5,fill="grey90")+
  geom_line(aes(x = timestamp, y = hr.bpm))+
  ggtitle(paste(day, bat))
    theme_cowplot()

p.tb <- hr %>% filter(date(timestamp) == day & tagID == "1UHL1865") %>% 
  ggplot()+
  geom_rect(xmin = ymd_hms(paste(day, "00:00:00"), tz= "America/Panama"),
            xmax = ymd_hms(paste(day, "06:00:00"), tz= "America/Panama"), ymin=-Inf, ymax=Inf, alpha=0.5,fill="grey90")+
  geom_rect(xmin = ymd_hms(paste(day, "19:00:00"), tz= "America/Panama"),
            xmax = ymd_hms(paste(day, "24:00:00"), tz= "America/Panama"), ymin=-Inf, ymax=Inf, alpha=0.5,fill="grey90")+
  geom_line(aes(x = timestamp, y = temp), color = "red")+
    theme_cowplot()

plot_grid(p.hr, p.tb, ncol = 1)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
